Energy Clustering

### 2017-09-17 qcqp_clustering/draft/energy_clustering.tex
continue commenting

\title{Nonparametric Clustering from  Energy Statistics}

@@

% @jv: stopped here
Based on Proposition~\ref{th:qcqp2}, to group data
$\mathbb{X} = \{ x_1,\dotsc,x_n \}$
into  $k$ clusters we first compute the Gram matrix
$G$ and then 
solve the optimization problem \eqref{eq:qcqp2} for $\Zt \in
\mathbb{R}^{n\times k}$. The $i$th row
of $\Zt$ will contain a single nonzero element in some $j$th column,
indicating that $x_i \in \C_j$. 
Problem \eqref{eq:qcqp2} is NP-hard and there
are few methods
available to solve it directly,
which is computational prohibitive even for small datasets.
However, one can find approximate solutions by relaxing some 
of the constraints, or obtaining a relaxed SDP version of it.
For instance, the relaxed problem
\begin{equation}
\max_{Y} \Tr \left( Y^\top G \, Y \right) \quad \mbox{s.t. $Y^\top Y = I$}
\end{equation}
has a well-known closed form solution $Y^\star = U R$, where the
columns of $U$ contain the leading $k$ eigenvectors of $G$ corresponding
to the $k$ largest eigenvalues $\lambda_1\ge \lambda_2\ge\dotsc\ge\lambda_k$, 
and
$R \in \mathbb{R}^{k\times k}$ is an arbitrary orthogonal matrix. 
The resulting
optimal objective function is given by
$\max \Tr \left( {Y^\star}^\top G \, Y^\star \right)  = 
\sum_{i=1}^k \lambda_i$. One might then normalize and threshold the rows
of $Y^\star$, or better, following \cite{NgJordan} one can normalize the
rows of $Y^\star$ and apply standard $k$-means on this matrix where each
row is considered as a data point.
This is the procedure done in spectral clustering on a graph Laplacian
matrix obtained through a similarity matrix.
However, computing eigenvectors of very large matrices
can be prohibitively expensive 
and usually iterative methods are preferred. Moreover,
this procedure is only guaranteed to converge if the normalized $Y^\star$ is
positive semidefinite.
In Section~\ref{sec:algo} we will propose an iterative method to directly 
find approximate solutions to \eqref{eq:qcqp2} that is guaranteed to
converge and does not require that the similarities $\kk(x_i,x_j)$ be
nonnegative.

% @jovo: stopped here.

@@

The main goal of the 
experiments  to follow
is to compare Algorithm~\ref{algo} based on Hartigan's method to
kernel $k$-means, as described in Algorithm~\ref{kmeans_algo}, which
is based on Lloyd's method. 
From the discussion in the end of the previous section we can anticipate
that Algorithm~\ref{algo} will be superior than Algorithm~\ref{kmeans_algo}.
Another goal is to illustrate the nonparametric aspect of energy statistics.
To this end we also compare
Algorithm~\ref{algo} to standard $k$-means and GMM/EM since 
these are reference
clustering algorithms in practice.
Moreover, for every algorithm, we always
choose the initialization procedure from $k$-means++%
\footnote{Notice that we just use the initialization
procedure and not the full $k$-means++ algorithm.} \cite{Vassilvitskii}.
Our measure of clustering quality will be the accuracy \eqref{eq:accuracy}
based on the ground truth. Furthermore, in every experiment, we sample
data many times and show the average value of the accuracy with error bars
indicating the standard error. Whenever possible, 
% @gui: bayes error is always possible to numerically estimate, no? maybe sometimes computationally intractable?
we also indicate the optimal 
accuracy computed from
Bayes classification error.

From the results of \cite{Telgarsky}, we expect the improvement of Hartigan's 
over Lloyd's method to be more accentuated in high dimensions.
Thus, we analyze
how the algorithms degrade as the number of dimensions increase, while
keeping the number of points in each cluster fixed. Consider
two clusters with multivariate normal distributions given by
\begin{equation}
\label{eq:gauss1}
\begin{split}
x &\in \C_i  \sim 
\mathcal{N}(\mu_i,\Sigma_i) \qquad (i=1,2),  \\
\mu_1 &= (\underbrace{0,\dotsc,0}_{\times D})^\top , \quad
\mu_2 = 0.7 \times (\underbrace{1,\dots,1}_{\times 10},
\underbrace{0,\dots,0}_{\times (D-10)})^\top, \quad
\Sigma_1 = \Sigma_2 = I_D.
\end{split}
\end{equation}
Note that the Bayes error
is fixed as $D$ increases, giving an optimal 
accuracy of $\approx 0.86$.
For each $D$ we sample $100$ times, picking $100$ points for each cluster.
We apply each algorithm to the resulting dataset 
and compute the average of the accuracy \eqref{eq:accuracy} (error bars
are standard error).
Algorithm~\ref{algo} and Algorithm~\ref{kmeans_algo} 
both use the standard
kernel \eqref{eq:standard_metric}.
The results are shown in Fig.~\ref{fig:gauss}a.
Note that GMM/EM is unable to estimate the covariance matrices
when the number of dimensions exceeds the number of points in each cluster,
giving zero accuracy when dimensions are $\gtrsim 100$.
We see that Algorithm~\ref{algo} performs better than all the other ones,
and in particular it outperforms kernel $k$-means algorithm as the number
of dimensions increase. Note also that although GMM and $k$-means are
parametric they are consistent models to the data.

\begin{figure}
\begin{minipage}{0.49\textwidth}
\includegraphics[width=1\textwidth]{gauss_dim.pdf}\\[-1.0em] (a)
\end{minipage}
\begin{minipage}{0.49\textwidth}
\includegraphics[width=1\textwidth]{gauss_cov_squareroot.pdf}\\[-1.0em] (b)
\end{minipage}
\begin{minipage}{0.49\textwidth}
\includegraphics[width=1\textwidth]{gauss_cov_linear.pdf}\\[-1.0em] (c)
\end{minipage}
\begin{minipage}{0.49\textwidth}
\includegraphics[width=1\textwidth]{gauss_cov_square.pdf}\\[-1.0em] (d)
\end{minipage}
\caption{
\label{fig:gauss}
Comparison of Algorithm~\ref{algo}, Algorithm~\ref{kmeans_algo},
standard $k$-means and GMM/EM as the number of dimensions increase
in gaussian settings.
We compute the average of \eqref{eq:accuracy} over $100$ samples with
error bars being standard error. We have two clusters with $100$ points
each. 
(a) Data as in \eqref{eq:gauss1}, where the optimal accuracy
from Bayes error is the dashed line equal to $\approx 0.86$.
(b) Data from \eqref{eq:gauss2} with $q=1/2$ in \eqref{eq:cov}. 
(c) Data from \eqref{eq:gauss2} with $q=1$ in \eqref{eq:cov}.
(d) Data from \eqref{eq:gauss2} with $q=2$ in \eqref{eq:cov}.
The optimal accuracy from Bayes error in (b--d) is $\approx 1$.
}
\end{figure}

Let us now consider the same experiment as before with data distributed as
\begin{equation}
\label{eq:gauss2}
x \in \C_i  \sim 
\mathcal{N}(\mu_i,\Sigma_i) \quad (i=1,2),  \quad
\mu_1 = (\underbrace{0,\dotsc,0}_{\times D})^\top , \quad
\mu_2 = (\underbrace{1,\dots,1}_{\times 10},
\underbrace{0,\dots,0}_{\times (D-10)})^\top, 
\end{equation}
where
\begin{equation}
\label{eq:cov}
(\Sigma_1)_{ij} = \begin{cases}
i^{-q} \delta_{ij} & \mbox{if $i \le 10$} \\
\delta_{ij} & \mbox{if $10 < i \le D$}
\end{cases} \qquad
(\Sigma_2)_{ij} = \begin{cases}
i^{\,q} \delta_{ij} & \mbox{if $i \le 10$} \\
\delta_{ij} & \mbox{if $10 < i \le D$}
\end{cases} 
\end{equation}
and we choose $q \in \{ 1/2, 1, 2\}$. In these cases,
the best possible accuracy from Bayes classification error is $\approx 1$.
In Fig.~\ref{fig:gauss}b--d we have $q=1/2$, $q=1$, and $q=2$, respectivelly.
Again, GMM/EM is unable to estimate the covariance matrices
as dimensions get larger than $\gtrsim 100$, and it gives poor results even
for number of dimensions much lower than this. Note that GMM requires
a larger number of points to estimate the parameters accurately.
Algorithm~\ref{algo} outperforms
Algorithm~\ref{kmeans_algo}, and $k$-means degrades faster as 
$q$ increases.

>>>>

%\title{Nonparametric Clustering from  Energy Statistics}
\title{Energy Clustering}

@@

% @jv: stopped here
Based on Proposition~\ref{th:qcqp2}, to group data
$\mathbb{X} = \{ x_1,\dotsc,x_n \}$
into  $k$ clusters we first compute the Gram matrix
$G$ and then 
solve the optimization problem \eqref{eq:qcqp2} for $\Zt \in
\mathbb{R}^{n\times k}$. The $i$th row
of $\Zt$ will contain a single nonzero element in some $j$th column,
indicating that $x_i \in \C_j$. 
Problem \eqref{eq:qcqp2} is NP-hard and there
are few methods
available to solve it directly,
which is computational prohibitive even for small datasets.
However, one can find approximate solutions by relaxing some 
of the constraints, or obtaining a relaxed SDP version of it.
For instance, the relaxed problem
\begin{equation}
\max_{Y} \Tr \left( Y^\top G \, Y \right) \quad \mbox{s.t. $Y^\top Y = I$}
\end{equation}
has a well-known closed form solution $Y^\star = U R$, where the
columns of $U$ contain the leading $k$ eigenvectors of $G$ corresponding
to the $k$ largest eigenvalues $\lambda_1\ge \lambda_2\ge\dotsc\ge\lambda_k$, 
and
$R \in \mathbb{R}^{k\times k}$ is an arbitrary orthogonal matrix. 
The resulting
optimal objective function is given by
$\max \Tr \left( {Y^\star}^\top G \, Y^\star \right)  = 
\sum_{i=1}^k \lambda_i$. One might then normalize and threshold the rows
of $Y^\star$,
% @gui: this might be a better initialization than k-means++, which essentially initializes using a different kernel?
 or better, following \cite{NgJordan} one can normalize the
rows of $Y^\star$ and apply standard $k$-means on this matrix where each
row is considered as a data point.
This is the procedure done in spectral clustering on a graph Laplacian
matrix obtained through a similarity matrix.
However, computing eigenvectors of very large matrices
can be prohibitively expensive 
and usually iterative methods are preferred. Moreover,
this procedure is only guaranteed to converge if the normalized $Y^\star$ is
positive semidefinite.
In Section~\ref{sec:algo} we will propose an iterative method to directly 
find approximate solutions to \eqref{eq:qcqp2} that is guaranteed to
converge and does not require that the similarities $\kk(x_i,x_j)$ be
nonnegative.

@@

The main goal of the 
experiments  to follow
is to compare Algorithm~\ref{algo} based on Hartigan's method to
kernel $k$-means, as described in Algorithm~\ref{kmeans_algo}, which
is based on Lloyd's method. 
% @gui: i might call them \mathcal{E}-H & \mathcal{E}-L, or some such?
From the discussion in the end of the previous section we can anticipate
that Algorithm~\ref{algo} will be superior than Algorithm~\ref{kmeans_algo}.
Another goal is to illustrate the nonparametric aspect of energy statistics.
To this end we also compare
Algorithm~\ref{algo} to standard $k$-means and GMM/EM since 
these are reference
clustering algorithms in practice.
Moreover, for every algorithm, we always
choose the initialization procedure from $k$-means++%
\footnote{Notice that we just use the initialization
procedure and not the full $k$-means++ algorithm.} \cite{Vassilvitskii}.
Our measure of clustering quality will be the accuracy \eqref{eq:accuracy}
based on the ground truth. Furthermore, in every experiment, we sample
data many times and show the average value of the accuracy with error bars
indicating the standard error. Whenever possible, 
% @gui: bayes error is always possible to numerically estimate, no? maybe sometimes computationally intractable?
we also indicate the optimal 
accuracy computed from
Bayes classification error.

From the results of \cite{Telgarsky}, we expect the improvement of Hartigan's 
over Lloyd's method to be more accentuated in high dimensions.
Thus, we analyze
how the algorithms degrade as the number of dimensions increase, while
keeping the number of points in each cluster fixed. Consider
two clusters with multivariate normal distributions given by
\begin{equation}
\label{eq:gauss1}
\begin{split}
x &\in \C_i  \sim 
\mathcal{N}(\mu_i,\Sigma_i) \qquad (i=1,2),  \\
\mu_1 &= (\underbrace{0,\dotsc,0}_{\times D})^\top , \quad
\mu_2 = 0.7 \times (\underbrace{1,\dots,1}_{\times 10},
\underbrace{0,\dots,0}_{\times (D-10)})^\top, \quad
\Sigma_1 = \Sigma_2 = I_D.
\end{split}
\end{equation}
Note that the Bayes error
is fixed as $D$ increases, giving an optimal 
accuracy of $\approx 0.86$.
For each $D$ we generate $100$ Monte Carlo runs, sampling $100$ points for each cluster.
% @gui: if you sample 100 points from each cluster, it breaks the GMM model, which assumes you sample n_i points from cluster i with probability \pi_i.  i don't think it matters that much, but good to know :)
We apply each algorithm to the resulting dataset 
and compute the average of the accuracy \eqref{eq:accuracy} (error bars
are standard error).
Algorithm~\ref{algo} and Algorithm~\ref{kmeans_algo} 
both use the standard
kernel \eqref{eq:standard_metric}.
The results are shown in Fig.~\ref{fig:gauss}a.
Note that GMM/EM is unable to estimate the covariance matrices
when the number of dimensions exceeds the number of points in each cluster,
giving zero accuracy when dimensions are $\gtrsim 100$.
We see that Algorithm~\ref{algo} performs better than all the other ones,
and in particular it outperforms kernel $k$-means algorithm as the number
of dimensions increase. 
% @gui: when you say "kernel k-means" here, do you mean algorithm 2?  if so, please clarify, if not, please clarify :)
Note also that although GMM and $k$-means are
parametric they are consistent models to the data.
% @guiL gmm & k-means are not 'models". they are algorithms, or in this context, they are estimators.

\begin{figure}
\begin{minipage}{0.49\textwidth}
\includegraphics[width=1\textwidth]{gauss_dim.pdf}\\[-1.0em] (a)
\end{minipage}
\begin{minipage}{0.49\textwidth}
\includegraphics[width=1\textwidth]{gauss_cov_squareroot.pdf}\\[-1.0em] (b)
\end{minipage}
\begin{minipage}{0.49\textwidth}
\includegraphics[width=1\textwidth]{gauss_cov_linear.pdf}\\[-1.0em] (c)
\end{minipage}
\begin{minipage}{0.49\textwidth}
\includegraphics[width=1\textwidth]{gauss_cov_square.pdf}\\[-1.0em] (d)
\end{minipage}
\caption{
\label{fig:gauss}
Comparison of Algorithm~\ref{algo}, Algorithm~\ref{kmeans_algo},
standard $k$-means and GMM/EM as the number of dimensions increase
in gaussian settings.
We compute the average of \eqref{eq:accuracy} over $100$ samples with
error bars being standard error. We have two clusters with $100$ points
each. 
(a) Data as in \eqref{eq:gauss1}, where the optimal accuracy
from Bayes error is the dashed line equal to $\approx 0.86$.
(b) Data from \eqref{eq:gauss2} with $q=1/2$ in \eqref{eq:cov}. 
(c) Data from \eqref{eq:gauss2} with $q=1$ in \eqref{eq:cov}.
(d) Data from \eqref{eq:gauss2} with $q=2$ in \eqref{eq:cov}.
The optimal accuracy from Bayes error in (b--d) is $\approx 1$.
% @gui: titles for each, or subcaptions, would be extremely useful.  in general, titles should provide the "setting" of the experiment, ie, the model, etc.
% @gui: also helpful here would be pairs plots, say, of the first 5 dimensions, showing the performance of our stuff (color points by Alg 3, maybe symbols denote "truth".  i have no intuition how difficult these are.
}
\end{figure}

Let us now consider the same experiment as before with data distributed as
% @gui: what do you mean "the same experiment as before"? clarify, or don't say that. there are obviously differences. 
\begin{equation}
\label{eq:gauss2}
x \in \C_i  \sim 
\mathcal{N}(\mu_i,\Sigma_i) \quad (i=1,2),  \quad
\mu_1 = (\underbrace{0,\dotsc,0}_{\times D})^\top , \quad
\mu_2 = (\underbrace{1,\dots,1}_{\times 10},
\underbrace{0,\dots,0}_{\times (D-10)})^\top, 
\end{equation}
where
\begin{equation}
\label{eq:cov}
(\Sigma_1)_{ij} = \begin{cases}
i^{-q} \delta_{ij} & \mbox{if $i \le 10$} \\
\delta_{ij} & \mbox{if $10 < i \le D$}
\end{cases} \qquad
(\Sigma_2)_{ij} = \begin{cases}
i^{\,q} \delta_{ij} & \mbox{if $i \le 10$} \\
\delta_{ij} & \mbox{if $10 < i \le D$}
\end{cases} 
\end{equation}
and we choose $q \in \{ 1/2, 1, 2\}$. In these cases,
% @gui: what is delta_ij??? can you explain the intuition for this setting, i don't get it, maybe because i don't know what delta_ij is.....
the best possible accuracy from Bayes classification error is $\approx 1$.
In Fig.~\ref{fig:gauss}b--d we have $q=1/2$, $q=1$, and $q=2$, respectivelly.
Again, GMM/EM is unable to estimate the covariance matrices
as dimensions get larger than $\gtrsim 100$, and it gives poor results even
for number of dimensions much lower than this. Note that GMM requires
a larger number of points to estimate the parameters accurately.
Algorithm~\ref{algo} outperforms
Algorithm~\ref{kmeans_algo}, and $k$-means degrades faster as 
$q$ increases.
